---
title: 'Random Variable Distribution Practice'
format: html
editor: source
---

# Practice Problems

## Using Wikipedia in college
A recent national study showed that approximately 44.7% of college students have used Wikipedia as a source in at least one of their term papers. Let $X$ equal the number of students in a random sample of size $n = 31$ who have used Wikipedia as a source.

a.  How is $X$ distributed?

::: {.callout-note collapse="true" title="Solution"}
**Assuming independence in sampling, and a representative sample, we can use a Binomial distribution with** $n=31$ and $p=0.447$.
:::

b.  Sketch the probability mass function (roughly).

::: {.callout-note collapse="true" title="Solution"}
```{r}
barplot(dbinom(0:31, 31,.447), names=0:31, ylab="probability", main="PMF of Binomial(31,.447)")
```
:::

c.  Sketch the cumulative distribution function (roughly).

::: {.callout-note collapse="true" title="Solution"}
```{r}
plot(pbinom(0:31, 31, .447), type="s", ylab="cumulative prob.", main="CDF of Binomial(31, .447)")
```
:::

d.  Find the probability that $X$ is equal to 17.

::: {.callout-note collapse="true" title="Solution"}
```{r}
dbinom(17, 31, .447)
```
:::

e.  Find the probability that $X$ is at most 13.

::: {.callout-note collapse="true" title="Solution"}
```{r}
pbinom(13, 31, .447)
```
:::

f.  Find the probability that $X$ is bigger than 11.

::: {.callout-note collapse="true" title="Solution"}
```{r}
sum(dbinom(12:31, 31, .447))
#or
pbinom(11, 31, .447, lower.tail=FALSE)
#or
1-pbinom(11, 31, .447)
```
:::

g.  Find the probability that $X$ is at least 15.

::: {.callout-note collapse="true" title="Solution"}
```{r}
#P(X at least 15)
sum(dbinom(15:31,31,.447))
```
:::

h.  Find the probability that $X$ is between 16 and 19, inclusive.

::: {.callout-note collapse="true" title="Solution"}
```{r}
sum(dbinom(16:19, 31, .447))
```
:::

i.  Give the mean of $X$, denoted $\mathbb{E}X$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
#E(X)=n*p
31*.447
#or you can also do this (but it's too much work)
sum( (0:31) * dbinom(0:31, 31, .447))

```
:::

j.  Give the variance of $X$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
#Var(X) = n * p * (1-p)
31 * .447 * (1-.447)
#or - if you want (but why would you want to?)
sum((0:31 - 31*.447)^2 * dbinom(0:31, 31, .447))
```
:::

k.  Give the standard deviation of $X$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
#SD(X) = sqrt(n*p*(1-p))
sqrt(31*.447*(1-.447))
```
:::

l.  Find $\mathbb{E}(4X+51.324)$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
#E(4X+51.324) = 4*E(X)+51.324
4*(31*.447) + 51.324
```
:::

## A Uniform PMF
Let $X$ have discrete uniform PMF on the values $x\in\left\{-1,0,1\right\}$. 

(a) Write the equation for its PMF.

(b) Find $Pr[X<-1]$ and $Pr[X\leq -1]$

(c) Find $Pr[X>0]$ and $Pr[X \geq 0]$

(d) Calculate the CDF from the PDF. Write out an expression for $F(x)$ and plot the PMF and CDF.

## Discrete Random Variable Problems

<!-- -->

a.  If $X$ is $\text{Poisson}(\lambda)$, compute $\mathbb{E}\left[1/(X+1)\right]$.

::: {.callout-note collapse="true" title="Solution"}
**This can be handled mathematically. The formula for** $E(1/(X+1))$ **is**

$E(1/(X+1))=\sum_{x=0}^{\infty}\frac{1}{x+1}\frac{\lambda^{x}}{x!}e^{-\lambda}=\sum_{x=0}^{\infty}\frac{\lambda^{x}}{(x+1)!}e^{-\lambda}$

**The trick is to get get the summation to equal 1 and simplify. We multiply by** $\lambda/\lambda$

$E(1/(X+1))=\frac{1}{\lambda}\sum_{x=0}^{\infty}\frac{\lambda^{x+1}}{(x+1)!}e^{-\lambda}$

**Now we can make a change of variables:** $y=x+1$ **and thus** $x=0$ **becomes** $y=1$

$E(1/(X+1)) = \frac{1}{\lambda}\sum_{y=1}^{\infty}\frac{\lambda^{y}}{y!}e^{-\lambda}$

**The only thing missing is that the summation starts at** $y=1$ **instead of** $y=0$**, But for** $Y \sim Poisson(\lambda)$**,** $P(Y=0)=e^{-\lambda}$ **so this summation is** $1-e^{-\lambda}$**.**

$E(1/(X+1)) = \frac{1}{\lambda}(1-e^{-\lambda})$
:::

b.  If $X$ is $\text{Bernoulli}(p)$ and $Y$ is $\text{Bernoulli}(q)$, computer $\mathbb{E}\left[(X+Y)^3\right]$ assuming $X$ and $Y$ are independent.

::: {.callout-note collapse="true" title="Solution"}
$(X+Y)^3 = X^3+3X^2Y+3XY^2+Y^3$ **so** $E[(X+Y)^3]=E(X^3)+3E(X^2)E(Y)+3E(X)E(Y^2)+E(Y^2)$

**this is due to independence. Since** $X$ **an** $Y$ **are independent, so are** $X^2$ **and** $Y$**, and** $X$ **and** $Y^2$**.** $E(X)=E(X^2)=E(X^3)=p$ **and** $E(Y)=E(Y^2)=E(Y^3)=q$**. Thus** $E[(X+Y)^3]=p+6pq+q$
:::

c.  Let $X$ be a random variable with mean $\mu$ and variance $\sigma^2$. Let $\Delta(\theta)=\mathbb{E}\left[(X-\theta)^2\right]$. Find $\theta$ that minimizes the error $\Delta(\theta)$.

::: {.callout-note collapse="true" title="Solution"}
**We can expand the expected value and attempt to find the minimum with respect to** $\theta$**.** $E[(X-\theta)^2]=E[X^2-2\theta X+\theta^2]=E(X^2)-2\theta\mu+\theta^2$**. Recall that** $Var(X)=E(X^2)-\mu^2$ **so** $E(X^2)=\sigma^2+\mu^2$ **So we can write** $\Delta(\theta)=\sigma^2 + \mu^2-2\theta\mu + \theta^2$ **We want to find what value of** $\theta$ **minimizes this function - derivative!** $\Delta'(\theta)=-2\mu+2\theta=0$ **thus** $\theta=\mu$ **minimizes this.**
:::

d.  Suppose that $X_1, \ldots, X_n$ are independent uniform random variables in $\{0,1,\ldots,100\}$. Evaluate $\mathbb{P}\left[\text{min}(X_1,\ldots, X_n) > l\right]$ for any $l \in \{0,1,\ldots,100\}$.

::: {.callout-note collapse="true" title="Solution"}
**Let** $Y=\min(X_1, \ldots, X_n)$ **If** $P(Y >l)$**, that means the minimum exceeds** $l$**, so all of the values** $>l$**.** $P(X_1 > l)=(100-l)/101$ **- you can check:** $P(X_1>0)=100/101$**. This is the same calculation for each** $i$**. So** $P(Y>l)=\dfrac{(100-l)^n}{101^n}$**.**
:::

e.  Consider a binomial random variable $X$ with parameters $n$ and $p$. $p_X(k)={n \choose k} p^k(1-p)^{n-k}$. Show that the mean is $\mathbb{E}X= np$.

::: {.callout-note collapse="true" title="Solution"}
$E(X)=\sum_{k=0}^n k{n \choose k} p^k(1-p)^{n-k}$

**The first term is zero so we could write**

$\sum_{k=1}^n k{n \choose k} p^k(1-p)^{n-k}$

**Now the following is a fact that is needed but perhaps not well known. It's the equivalence of** $k{n \choose k}=n{n-1 \choose k-1}$**. We make this subsitution**

$\sum_{k=1}^n n{n-1 \choose k-1} p^k(1-p)^{n-k}=np\sum_{k=1}^n {n-1\choose k-1}p^{k-1}(1-p)^{n-k}$

**We could write** $n-k=(n-1)-(k-1)$ **and we'll be making some substitutions:** $m=n-1$ **and** $j=k-1$**. This lets us write**

$np\sum_{j=0}^m {m \choose j}p^j(1-p)^{m-j}=np$ **because the summation =1, as it is just the sum of the pmf of a binomial.**
:::

f.  **(not for 340)** Consider a geometric random variable $X$ with parameter $p$. $p_X(k)=p(1-p)^k$ for $k=0,1,\ldots$. Show that its mean is $\mathbb{E}X=(1-p)/p$.

g.  **(not for 340)** Consider a Poisson random variable $X$ with parameter $\lambda$. $p_X(k)=\dfrac{\lambda^k}{k!}e^{-\lambda}$. Show that $\text{Var}X=\lambda$.

h.  **(not for 340)** Consider the uniform random variable $X$ over values $1,2,\ldots, L$. Show that $\text{Var}X=\dfrac{L^2-1}{12}$. *Hint:* $\sum_{i=1}^n i = \frac{n(n+1)}{2}$ and $\sum_{i=1}^n i^2=\frac{n^3}{3}+\frac{n^2}{2}+\frac{n}{6}$

<!-- -->

## Hard Drive Failures
An audio player uses a low-quality hard drive. The probability that the hard drive fails after being used for one month is 1/12. If it fails, the manufacturer offers a free-of-charge repair for the customer. For the cost of each repair, however, the manufacturer has to pay \$20. The initial cost of building the player is \$50, and the manufacturer offers a 1-year warranty. Within one year, the customer can ask for a free repair up to 12 times.

<!-- -->

a.  Let $X$ be the number of months when the player fails. What is the PMF of $X$? *Hint:* $\mathbb{P}(X = 1)$ *may not be very high because if the hard drive fails it will be fixed by the manufacturer. Once fixed, the drive can fail again in the remaining months. So saying* $X = 1$ *is equivalent to saying that there is only one failure in the entire 12-month period.*

::: {.callout-note collapse="true" title="Solution"}
**The number of failures should follow a binomial distribution with** $n=12, p=1/12$**. Thus** $P(X=k)={n \choose k}(\frac{1}{12})^k(\frac{11}{12})^{n-k}$
:::

b.  What is the average cost per player?

::: {.callout-note collapse="true" title="Solution"}
**The cost is** $50+20X$ **So** $E(50+20X)=50+20E(X)=50+20\cdot 12(\frac{1}{12})=70$
:::

## Bit Errors
15. A binary communication channel has a probability of bit error of $p = 10^{-6}$. Suppose that transmission occurs in blocks of 10,000 bits. Let $N$ be the number of errors introduced by the channel in a transmission block.

<!-- -->

a.  What is the PMF of $N$?

::: {.callout-note collapse="true" title="Solution"}
$N$ **follows a binomial distribution with** $n=10000$ **and** $p=.000001$
:::

b.  Find $\mathbb{P}(N = 0)$ and \$\mathbb{P}(N b \$ 3)\$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
dbinom(0, 10000, .000001)
pbinom(3, 10000, .000001)
```
:::

c.  For what value of $p$ will the probability of 1 or more errors in a block be 99%?

::: {.callout-note collapse="true" title="Solution"}
**This can be solved directly.** $P(N \geq 1)=1-P(X=0)=1-(1-p)^{10000}$**. If we set this to .99 we can solve for** $p$ **:** $.99=1-(1-p)^{10000}$ **so** $.01 = (1-p)^{10000}$ **so** $p=1-.01^{1/10000}$

```{r}
1-.01^(1/10000)
```
:::

## Processing Orders
The number of orders waiting to be processed is given by a Poisson random variable with parameter $\alpha = \frac{\lambda}{n\mu}$, where $\lambda$ is the average number of orders that arrive in a day, $\mu$ is the number of orders that an employee can process per day, and n is the number of employees. Let $N; = 5$ and $N< = 1$. Find the number of employees required so the probability that more than four orders are waiting is less than 10%.

Hint: You need to use trial and error for a few $n$'s.

::: {.callout-note collapse="true" title="Solution"}
```{r}
lambda=5
mu=1
ppois(4, lambda/(1:10 * mu), lower.tail=FALSE)
#With 3 employees P(X>4) is less than 10%.
```
:::

## Normal Random Variable
If $Z\sim \text{Normal}(\mu=0, \sigma^2=1^2)$ find

<!-- -->

a.  $\mathbb{P}(Z > 2.64)$

::: {.callout-note collapse="true" title="Solution"}
```{r}
pnorm(2.64, 0, 1, lower.tail=FALSE)
```
:::

b.  $\mathbb{P}(0 \leq Z < 0.87)$

::: {.callout-note collapse="true" title="Solution"}
```{r}
pnorm(.87)-pnorm(0)
```
:::

c.  $\mathbb{P}(|Z| > 1.39)$ (*Hint:* draw a picture)

::: {.callout-note collapse="true" title="Solution"}
```{r}
pnorm(1.39, lower.tail=FALSE)*2
```
:::

## Identify the Distribution
For the following random experiments, decide what the distribution of X should be. In nearly every case, there are additional assumptions that should be made for the distribution to apply; identify those assumptions (which may or may not strictly hold in practice).

<!-- -->

a.  We throw a dart at a dart board. Let X denote the squared linear distance from the bullseye to the where the dart landed.

::: {.callout-note collapse="true" title="Solution"}
**Assume the dart lands somewhere on the board, and any point is equally likely (not a good assumption for a skilled dart thrower). The probability density would be proportional to the distance to the center squared - Suppose the dart board has radius** $R$. **Let** $X$ **be the distance to the dart from the bullseye. Then** $P(X<r)=\pi r^2 / (\pi R^2)=(r/R)^2$ **. The question then is what is** $P(X^2<r)$**? Well, take a square root of both sides.** $=P(X < \sqrt{r})=\frac{r}{R^2}$**. This is a uniform distribution's CDF.**
:::

b.  We randomly choose a textbook from the shelf at the bookstore and let P denote the proportion of the total pages of the book devoted to exercises.

::: {.callout-note collapse="true" title="Solution"}
**A random proportion you might want to use uniform(0,1) however this is assuming that each proportion is equally likely. This is actually a great example for a beta distribution. Beta distributions are continuous distributions that can be parameterized to model a random proportion and the distribution can can be made to be skewed in different ways.**
:::

c.  We measure the time it takes for the water to completely drain out of the kitchen sink.

::: {.callout-note collapse="true" title="Solution"}
**Let's assume the sink is filled to the maximum. We drain the sink and start our timer. In this case, it's reasonable to model the length of time to drain as a normal distribution.**
:::

d.  We randomly sample strangers at the grocery store and ask them how long it will take them to drive home.

::: {.callout-note collapse="true" title="Solution"}
**The time it takes to go home could be modeled by a gamma distribution since it is a continuous distribution capped below at 0 and it is a useful way to model the length of time a random process takes to complete.**
:::

## Normal Random Variable II
Let $X$ be a Gaussian random variable with $\mu=5$ and $\sigma^2=16$.

<!-- -->

a.  Find $\mathbb{P}(X>4)$ and $\mathbb{P}(2\leq X \leq 7)$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
#P(X>4)
pnorm(4, mean=5, sd=4, lower.tail=FALSE)

#P(2 <= X <= 7)
pnorm(7, 5, 4)-pnorm(4,5,4)
```
:::

b.  If $\mathbb{P}(X < a)=0.8869$, find $a$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
qnorm(.88695, 4)
```
:::

c.  If $\mathbb{P}(X>b)=0.1131$, find $b$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
qnorm(.1131, 5, 4, lower.tail=FALSE)
```
:::

d.  If $\mathbb{P}(13 < X \leq c)=0.0011$, find $c$.

::: {.callout-note collapse="true" title="Solution"}
```{r}
#First find the probability less than 13
p13 <- pnorm(13, 5, 4)
#now we can find the quantile for p13+.0011
qnorm(p13+.0011, 5, 4)

#double check
pnorm(13.08321,5,4)-pnorm(13,5,4)
```
:::

## Choose the distribution
2.  For the following situations, decide what the distribution of $X$ should be. In nearly every case, there are additional assumptions that should be made for the distribution to apply; identify those assumptions (which may or may not hold in practice.)

a.  We shoot basketballs at a basketball hoop, and count the number of shots until we make a basket. Let X denote the number of missed shots. On a normal day we would typically make about 37% of the shots.

::: {.callout-note collapse="true" title="Solution"}
**The number of missed shots before the first basket, assuming independence, can be modeled by a Geometric random variable with parameter** $p=.37$**.**
:::

b.  In a local lottery in which a three digit number is selected randomly, let X be the number selected.

::: {.callout-note collapse="true" title="Solution"}
**Assuming that all 3 digit numbers are equally likely (A reasonable assumption) the number selected can be modeled by a discrete uniform distribution with minimum 100 and maximum 999.**
:::

c.  We drop a Styrofoam cup to the floor twenty times, each time recording whether the cup comes to rest perfectly right side up, or not. Let X be the number of times the cup lands perfectly right side up.

::: {.callout-note collapse="true" title="Solution"}
**If we drop the cup 20 times, and the result each time is independent with a constant probability of landing right side up, the number of times it does can be modeled by a Binomial random variable with parameters** $n=20$ **and** $p$ **(unknown).**
:::

d.  We toss a piece of trash at the garbage can from across the room. If we miss the trash can, we retrieve the trash and try again, continuing to toss until we make the shot. Let X denote the number of missed shots.

::: {.callout-note collapse="true" title="Solution"}
**Geometric random variable (unknown parameter value for** $p$**).**
:::

e.  Working for the border patrol, we inspect shipping cargo as when it enters the harbor looking for contraband. A certain ship comes to port with 557 cargo containers. Standard practice is to select 10 containers randomly and inspect each one very carefully, classifying it as either having contraband or not. Let X count the number of containers that illegally contain contraband.

::: {.callout-note collapse="true" title="Solution"}
**Technically we should use a hypergeometric random variable for this situation (since it is a small population size of 557), but since we do not cover the hypergeometric the closest random variable we have is the binomial.**
:::

f.  At the same time every year, some migratory birds land in a bush outside for a short rest. On a certain day, we look outside and let X denote the number of birds in the bush.

::: {.callout-note collapse="true" title="Solution"}
**This is a discrete random variable, but without other information it's hard to say. The distribution is likely unimodal and bell-curved. You could probably model this using a normal distribution rounded off to the nearest integer.**
:::

g.  We count the number of rain drops that fall in a circular area on a sidewalk during a ten minute period of a thunder storm.

::: {.callout-note collapse="true" title="Solution"}
**The observation window is the circular area, and the 10 minutes during observation. Assuming the rate of rainfall is constant, the number of raindrops in the circle can be modeled using a Poisson random variable.**
:::

h.  We count the number of moth eggs on our window screen.

::: {.callout-note collapse="true" title="Solution"}
**Counting indicates a discrete random variable. A binomial or a rounded normal distribution may be appropriate, but we lack enough details to be sure.**
:::

i.  We count the number of blades of grass in a one square foot patch of land.

::: {.callout-note collapse="true" title="Solution"}
**The location of the sprouting grass could be modeled well by a Poisson random variable - the** $\lambda$ **parameter would likely be very large, in the range of 1000 or 10000, and as such the distribution would look very much like a normal distribution.**
:::

j.  We count the number of pats on a baby's back until (s)he burps.

::: {.callout-note collapse="true" title="Solution"}
**As we define a geometric random variable, we let** $X$ **be the number of failures before the first success. The last pat (that causes the burp) is the success in this context. So we could use a geometric random variable, but we would have to add 1 to it in order to count all burps (the failures + 1 success).**
:::




## Two Normal RVs
Let X and Y be zero-mean, unit-variance independent Gaussian random variables. Find the value of r for which the probability that $(X, Y )$ falls inside a circle of radius r is 1/2.

::: {.callout-note collapse="true" title="Solution"}
```{r}
x <- rnorm(10000)
y <- rnorm(10000)

r <- seq(1.1, 1.2, .005)
p <- 0
for (i in 1:length(r)){
  p[i] <- mean(sqrt(x^2+y^2) <= r[i])
}
data.frame(r,p)


#X^2 + Y^2 ~ Chisq(2)
#so the square root of the 50th percentile from that distribution should be the answer
sqrt(qchisq(.5,2))

```
:::

## Uniform Random Angle
Let $\Theta ∼ Uniform[0, 2\pi]$.

<!-- -->

a.  If $X = cos \Theta$, $Y = sin \Theta$. Are $X$ and $Y$ uncorrelated?

::: {.callout-note collapse="true" title="Solution"}
**Yes, they are uncorrelated, because (x,y) can be any point on the circumference of a circle of radius 1 with uniform likelihood. However, they are not independent. If we know the value of** $Y$ for example, there are only 2 possible values of $X$.

```{r}
thetas <- runif(10000, 0, 2*pi)
cor(cos(thetas), sin(thetas))
```
:::

b.  If $X = cos(\Theta/4)$, $Y = sin(\Theta/4)$. Are $X$ and $Y$ uncorrelated?

::: {.callout-note collapse="true" title="Solution"}
**In this case (x,y) can only be found in the first quadrant. In this case they are going to be negatively correlated, since that portion of the unit circle in the first quadrant slopes downwards.**

```{r}
cor(cos(thetas/4), sin(thetas/4))
```
:::


# Beyond STAT 340

These problems are excellent practice but they are beyond the material we cover in STAT 340.


## Variance of a Uniform RV
Calculate the variance of $X \sim \text{Unif}(a,b)$. (*Hint:* First calculate $\mathbb{E}X^2$)

## Expectation and Binomial

If $X \sim \text{Binom}(n,p)$ show that $\mathbb{E}X(X-1)=n(n-1)p^2$.

::: {.callout-note collapse="true" title="Solution"}
**We can just expand the product** $\mathbb{E}(X^2-X)$ **and we can split this up into two expected values:** $\mathbb{E}X^2 - \mathbb{E}X = \mathbb{E}X^2-\mu$**. Recall that** $Var(X)=\mathbb{E}X^2-\mu^2$ **So** $\mathbb{E}X^2=Var(X)+\mu^2$**. For a binomial,** $Var(X)=np(1-p)$ **and** $\mu=np$**. Thus we have**

$\mathbb{E}X^2 - \mu=[np(1-p) + n^2p^2] - np = np\left(1-p+np-1\right)$

**Tidying up a little bit we get** $np(np-p)=np^2(n-1)$**, and we're done.**
:::


